Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-20 manuscript/main.tex
added pfg construction

%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[twocolumn]{ceurart}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2022}
\copyrightclause{
    Copyright for this paper by its authors.
    Use permitted under Creative Commons License Attribution 4.0 
    International (CC BY 4.0).
}

%%
%% This command is for the conference information
\conference{
    Woodstock'22: Symposium on the irreproducible science,
    June 07--11, 2022, Woodstock, NY
}

%%
%% The "title" command
\title{Prefix-free graphs and suffix array construction in sublinear space}

\tnotemark[1]
\tnotetext[1]{
    You can use this document as the template for preparing your
    publication. We recommend using the latest version of the ceurart style.
}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1,2]{Andrej Baláž}[%
orcid=0000-0002-0877-7063,
email=kulyabov-ds@rudn.ru,
url=https://yamadharma.github.io/,
]
\cormark[1]
\fnmark[1]
\address[1]{
    Peoples' Friendship University of Russia (RUDN University),
    6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation
}

%% Footnotes
\cortext[1]{Corresponding author.}
\fntext[1]{These authors contributed equally.}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  LaTeX class \sep
  paper template \sep
  paper formatting \sep
  CEUR-WS
\end{keywords}

@@

\section{Prefix-free graphs}
The idea of prefix-free graphs is inspired by

\section{Suffix array construction}
\section{Related work}
MARIA \cite{2022maria}
Prefix-free Wheeler graphs \cite{2022pfwg}
Prefix-free parsing for building big BWTs \cite{2019boucher}
Rpair: rescaling repair with rsync \cite{2019gagie}

@@

%%
%% The acknowledgments section is defined using the "acknowledgments" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acknowledgments}
  Thanks to the developers of ACM consolidated LaTeX styles
  \url{https://github.com/borisveytsman/acmart} and to the developers
  of Elsevier updated \LaTeX{} templates
  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.
\end{acknowledgments}

%%
%% Define the bibliography file to be used
\bibliography{sources}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

@@

The sources for the ceur-art style are available via
\begin{itemize}
\item \href{https://github.com/yamadharma/ceurart}{GitHub},
% \item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
\item
  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/pkfscdkgkhcq}{Overleaf
    template}.
\end{itemize}

>>>>

%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[twocolumn]{ceurart}

@@

\begin{document}

\copyrightyear{2023}
\copyrightclause{
    Copyright for this paper by its authors.
    Use permitted under Creative Commons License Attribution 4.0 
    International (CC BY 4.0).
}

\conference{
    ITAT'23: Information technologies -- Applications and Theory
    September 22--26, 2023, Tatranské Matliare, Vysoké Tatry
}

\title{Prefix-free graphs and suffix array construction in sublinear space}

\author[1]{Andrej Baláž}[
orcid=0000-0001-0002-0003,
email=andrejbalaz001@gmail.com,
]
\cormark[1]
\address[1]{
    Faculty of Mathematics, Physics and Informatics,
    Comenius University, Mlynská dolina F1, 842 48 Bratislava, Slovakia
}

% Footnotes
\cortext[1]{Corresponding author.}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  pangenome \sep
  suffix array
\end{keywords}

@@

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{}.
% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.
% add sketch of a proof

% how to construct prefix free graphs
To illustrate how to partition the sequences into segments consider a set of
strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
Next, assume we will partition the sequences using a set of trigger words 
$T = {AC, CG}$.

In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
In our case, we need to append $k$ sentinels to the sequences.
The reasons for doing so will become clear in the suffix array construction section.

Then, using a sliding window we can scan through the sequences and each time we
encounter a trigger word, we create a new segment from the start of the previous
trigger word to the end of the current trigger word.
Two special cases happen at the beginning of a sequence, when no previous
trigger was encountered, and at the end.
In the former case, the segment starts simply at the start of a sequence and in
the later case the segment ends at a sequence end.

During the sequence scan, each unique segment is assigned an ID.
The original sequence can then be represented as a list of IDs, also called 
\emph{path}s.

The result of this procedure is a set of segments and a list of paths.
In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
and the paths are $[  ]$.

Because each segment is represented in a set $S$ only once, the sum of lengths
of segments is usually much smaller than the length of original sequences.

This set of segments, together with the list of paths are already a prefix-free
graph, but in order to be useful for the next part, we need to normalize it.
By normalizing, we mean to sort the segments lexicographically and relabel the
paths accordingly.
This will give us a list of segments and a list of paths which can be directly
represented in the GFA format as shown in the Figure \ref{}.

The reconstruction of the original sequence can then be done by expanding 
the particular path, ignoring last $k$ characters of each segment.

\section{Suffix array construction}

\section{Related work}
MARIA \cite{2022maria}
Prefix-free Wheeler graphs \cite{2022pfwg}
Prefix-free parsing for building big BWTs \cite{2019boucher}
Rpair: rescaling repair with rsync \cite{2019gagie}

@@

\begin{acknowledgments}
    Thanks to Alessia and Askar for their encouragement while developing prefix-free graphs.
\end{acknowledgments}

\bibliography{sources}

\appendix

@@

The sources for the prefix-free graphs are available via
\begin{itemize}
\item \href{https://github.com/andynet/pfg}{GitHub}.
\end{itemize}
